Split-Apply-Combine with Dynamic Grouping

### 2024-03-27 paper/article.Rnw
updated indentation to JSS style

It is possible to get the same result with \code{accumulate()}. This will cause
summation over all variables that are not used in the formula object. In the
below example we also introduce the helper function \code{min_records()}.
<<>>=
input$Y2 <- 11:19
accumulate(input, collapse = A * B ~ A * B1 + A
          , test = min_records(3), fun = mean)
@
This means that for experimentation, users must be careful to exclude categorical
variables that are not used from the input data set. For example, if \code{B1}
is not used, we get the following.
<<>>=
accumulate(input[-3], collapse = A * B ~ A
          , test = min_records(3), fun = mean)
@

@@

In order to use the data frame interface, the input dataset must include the
most fine-grained grouping variable. In the running example this is $A\times
B$, so we need to combine that into a single variable, and remove the other
ones.
<<>>=
input1 <- input
input1$AB <- paste(input$A, input$B, sep = "-")
input1 <- input1[-(1:3)]
input1
@
We now define the collapsing scheme as follows.
<<>>=
csh <- data.frame(
  AB  = c("1-11", "2-12", "2-13", "3-21", "3-22", "3-12")
, AB1 = c("1-1" , "2-1" , "2-1" , "3-2" , "3-2" , "3-1" )
, A   = c("1"   , "2"   , "2"   , "3"   , "3"   , "3"   ))
csh
@
In this data frame, two consecutive columns should be read as a child-parent
relation. For example, in the first collapsing step the groups defined by
\code{AB == "2-12"} and \code{AB == "2-13"} both collapse to \code{AB1 == "2-1"}.
In this artificial example the codes do not mean anything, but in realistic
cases where codes represent a (hierarchical) classification, domain experts
usually have a good grasp of which codes can be combined.

The calls to \code{cumulate()} and \code{accumulate()} now look as follows.
<<>>=
accumulate(input1, collapse = csh, test = min_records(3), mean)
cumulate(input1, collapse = csh, test = min_records(3)
        , muY = mean(Y), muY2 = mean(Y2))
@

@@

\begin{center}
\begin{tikzpicture}
\tikzstyle{level 1}=[level distance=10mm,sibling distance=40mm]
\tikzstyle{level 2}=[level distance=10mm,sibling distance=10mm]
  \node {\code{01}}
    child { node {\code{011}}
      child {node {\code{0111}}}
      child {node {\code{0112}}} 
      child {node {\code{0113}}} }
    child { node {\code{012}}
      child {node {\code{0121}}}
      child {node {\code{0122}}} 
      child {node {\code{0123}}} 
      child {node {\code{0124}}} };
\end{tikzpicture}
\end{center}
Here, the hierarchy suggests to collapse $\{0111,0112,0113\}$ into $\{011\}$
when needed, and similarly for the right branch. The second level of collapsing
would combine $\{012\}$ with $\{011\}$ into $\{01\}$. The \pkg{accumulate}
package comes with a helper function that creates the collapsing scheme from the
lowest-level digits.
<<>>=
nace <- c("0111", "0112", "0113", "0121"
         , "0121", "0122", "0123", "0124")
csh_from_digits(nace, levels = 2)
@ 
Here, the parameter \code{levels} determines how many collapsing steps will be
computed. Since all codes are prepended with zero, there is no need to collapse
$01$ any further. The output can be used as argument to the \code{collapse}
parameter of the \code{accumulate()} or \code{cumulate()} functions.

@@

In the simplest case, where no auxiliary information is available or used, one
replaces the estimator of the mean over a subpopulation with the estimator of
the mean over a larger subpopulation that includes the target subpopulation
\citep[Section 3.2.1]{rao2015small}. If we assume that the dataset is obtained
by simple random sampling from the population, the mean can be estimated with
the sample mean. In this example we will demand that there are at least ten
records for which turnover has been measured. The collapsing scheme is given by
\code{sbi * size ~ sbi + sbi2 + sbi1} where \code{sbi2} and \code{sbi1} are
classification by respectively the first two SBI digits and the first SBI
digit. We first add those variables to the dataset.
<<>>=
producers$sbi2 <- substr(producers$sbi, 1, 2)
producers$sbi1 <- substr(producers$sbi, 1, 1)
head(producers, 3)
@
Using \code{cumulate()} we obtain the means. 
<<>>=
a <- cumulate(producers
             , collapse = sbi * size ~ sbi + sbi2 + sbi1
             , test = min_complete(n = 10, vars = "industrial")
             , mean_industrial = mean(industrial, na.rm = TRUE))
head(a,3)
@

@@

We use SAE to estimate the subpopulation ratios, and then use the \code{simputation}
package \citep{loo2022simputation} to impute the missing values. 
<<>>=
r <- cumulate(producers
         , collapse = sbi * size ~ sbi + sbi2 + sbi1
         , test = min_complete(n = 10, vars = "industrial")
         , R = mean(industrial, na.rm = TRUE)/mean(total, na.rm = TRUE))
head(r,3)
@
To impute the values, using \code{impute_proxy()} we need to merge the ratios
with the producers dataset (which automatically happens by SBI and size class).
<<>>=
library("simputation")
dat <- merge(producers, r)
dat <- impute_proxy(dat, industrial ~ R * total)
@
We can inspect the imputed values as follows.
<<>>=
iNA <- is.na(producers$industrial)
head(dat[iNA, c("sbi", "size", "level", "R", "industrial", "total")])
@
From this, we get all the information to interpret the imputed values. For
example, we see that in record 664, the ratio was estimated after collapsing
the group $(\texttt{sbi},\texttt{size})=(21122,7)$ to $\texttt{sbi2}=21$ since
the collapse \code{level} equals 2.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing Complex Aggregates} \label{sect:complex}
Until now, the aggregates have been simple (scalar) values.  With the
\code{cumulate()} function it is also possible to specify complex aggregates that
go beyond simple aggregates. Below, we estimate the following linear model
\begin{displaymath}
\texttt{total} = \beta_0 + \beta_i\texttt{industrial} + \varepsilon,
\end{displaymath}
demanding that there are at least 10 records where both predictor and predicted
variable are available.
<<>>=
r <- cumulate(producers
         , collapse = sbi * size ~ sbi + sbi2 + sbi1
         , test = min_complete(n = 10, vars = c("total","industrial"))
         , model = lm(total ~ industrial))
head(r,3)
@
Here, the last column is a list of class \code{object_list}, where each
element is either an object of class \code{lm} or \code{NA}. Variables of class
\code{object_list} only differ from from a standard \proglang{R} list by their
print method.

>>>>

It is possible to get the same result with \code{accumulate()}. This will cause
summation over all variables that are not used in the formula object. In the
below example we also introduce the helper function \code{min_records()}.
<<>>=
input$Y2 <- 11:19
accumulate(input, collapse = A * B ~ A * B1 + A, 
  test = min_records(3), fun = mean)
@
This means that for experimentation, users must be careful to exclude categorical
variables that are not used from the input data set. For example, if \code{B1}
is not used, we get the following.
<<>>=
accumulate(input[-3], collapse = A * B ~ A,
  test = min_records(3), fun = mean)
@

@@

In order to use the data frame interface, the input dataset must include the
most fine-grained grouping variable. In the running example this is $A\times
B$, so we need to combine that into a single variable, and remove the other
ones.
<<>>=
input1 <- input
input1$AB <- paste(input$A, input$B, sep = "-")
input1 <- input1[-(1:3)]
input1
@
We now define the collapsing scheme as follows.
<<>>=
csh <- data.frame(
  AB  = c("1-11", "2-12", "2-13", "3-21", "3-22", "3-12"),
  AB1 = c("1-1" , "2-1" , "2-1" , "3-2" , "3-2" , "3-1" ),
  A   = c("1"   , "2"   , "2"   , "3"   , "3"   , "3"   ))
csh
@
In this data frame, two consecutive columns should be read as a child-parent
relation. For example, in the first collapsing step the groups defined by
\code{AB == "2-12"} and \code{AB == "2-13"} both collapse to \code{AB1 == "2-1"}.
In this artificial example the codes do not mean anything, but in realistic
cases where codes represent a (hierarchical) classification, domain experts
usually have a good grasp of which codes can be combined.

The calls to \code{cumulate()} and \code{accumulate()} now look as follows.
<<>>=
accumulate(input1, collapse = csh, test = min_records(3), mean)
cumulate(input1, collapse = csh, test = min_records(3),
  muY = mean(Y), muY2 = mean(Y2))
@

@@

\begin{center}
\begin{tikzpicture}
\tikzstyle{level 1}=[level distance=10mm,sibling distance=40mm]
\tikzstyle{level 2}=[level distance=10mm,sibling distance=10mm]
  \node {\code{01}}
    child { node {\code{011}}
      child {node {\code{0111}}}
      child {node {\code{0112}}} 
      child {node {\code{0113}}} }
    child { node {\code{012}}
      child {node {\code{0121}}}
      child {node {\code{0122}}} 
      child {node {\code{0123}}} 
      child {node {\code{0124}}} };
\end{tikzpicture}
\end{center}
Here, the hierarchy suggests to collapse $\{0111,0112,0113\}$ into $\{011\}$
when needed, and similarly for the right branch. The second level of collapsing
would combine $\{012\}$ with $\{011\}$ into $\{01\}$. The \pkg{accumulate}
package comes with a helper function that creates the collapsing scheme from the
lowest-level digits.
<<>>=
nace <- c("0111", "0112", "0113", "0121", "0121", "0122", "0123", "0124")
csh_from_digits(nace, levels = 2)
@ 
Here, the parameter \code{levels} determines how many collapsing steps will be
computed. Since all codes are prepended with zero, there is no need to collapse
$01$ any further. The output can be used as argument to the \code{collapse}
parameter of the \code{accumulate()} or \code{cumulate()} functions.

@@

In the simplest case, where no auxiliary information is available or used, one
replaces the estimator of the mean over a subpopulation with the estimator of
the mean over a larger subpopulation that includes the target subpopulation
\citep[Section 3.2.1]{rao2015small}. If we assume that the dataset is obtained
by simple random sampling from the population, the mean can be estimated with
the sample mean. In this example we will demand that there are at least ten
records for which turnover has been measured. The collapsing scheme is given by
\code{sbi * size ~ sbi + sbi2 + sbi1} where \code{sbi2} and \code{sbi1} are
classification by respectively the first two SBI digits and the first SBI
digit. We first add those variables to the dataset.
<<>>=
producers$sbi2 <- substr(producers$sbi, 1, 2)
producers$sbi1 <- substr(producers$sbi, 1, 1)
head(producers, 3)
@
Using \code{cumulate()} we obtain the means. 
<<>>=
a <- cumulate(producers,
  collapse = sbi * size ~ sbi + sbi2 + sbi1,
  test = min_complete(n = 10, vars = "industrial"),
  mean_industrial = mean(industrial, na.rm = TRUE))
head(a,3)
@

@@

We use SAE to estimate the subpopulation ratios, and then use the \code{simputation}
package \citep{loo2022simputation} to impute the missing values. 
<<>>=
r <- cumulate(producers,
  collapse = sbi * size ~ sbi + sbi2 + sbi1,
  test = min_complete(n = 10, vars = "industrial"),
  R = mean(industrial, na.rm = TRUE)/mean(total, na.rm = TRUE))
head(r,3)
@
To impute the values, using \code{impute_proxy()} we need to merge the ratios
with the producers dataset (which automatically happens by SBI and size class).
<<>>=
library("simputation")
dat <- merge(producers, r)
dat <- impute_proxy(dat, industrial ~ R * total)
@
We can inspect the imputed values as follows.
<<>>=
iNA <- is.na(producers$industrial)
head(dat[iNA, c("sbi", "size", "level", "R", "industrial", "total")])
@
From this, we get all the information to interpret the imputed values. For
example, we see that in record 664, the ratio was estimated after collapsing
the group $(\texttt{sbi},\texttt{size})=(21122,7)$ to $\texttt{sbi2}=21$ since
the collapse \code{level} equals 2.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computing Complex Aggregates} \label{sect:complex}
Until now, the aggregates have been simple (scalar) values.  With the
\code{cumulate()} function it is also possible to specify complex aggregates that
go beyond simple aggregates. Below, we estimate the following linear model
\begin{displaymath}
\texttt{total} = \beta_0 + \beta_i\texttt{industrial} + \varepsilon,
\end{displaymath}
demanding that there are at least 10 records where both predictor and predicted
variable are available.
<<>>=
r <- cumulate(producers,
  collapse = sbi * size ~ sbi + sbi2 + sbi1,
  test = min_complete(n = 10, vars = c("total","industrial")),
  model = lm(total ~ industrial))
head(r,3)
@
Here, the last column is a list of class \code{object_list}, where each
element is either an object of class \code{lm} or \code{NA}. Variables of class
\code{object_list} only differ from from a standard \proglang{R} list by their
print method.
